Generating Empirically Optimized Numerical Software from MATLAB Prototypes

### 2008-09-19 doc/papers/cgo08/paper.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@349 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

% CGO 09: http://www.cgo.org/
% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
% Max 22 pages, double-spaced, 11pt or larger font
% Abstract: Sep. 12, 2008
% Full paper: Sep. 19, 2008
\documentclass[11pt]{article}

@@

\abstract{
The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology. Programming scientific applications is hard, and optimizing them for high performance is even harder.  The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations. Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.  Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.

@@

%Words here in Intro and Back are lifted from a 2004 proposal by Liz, Bill, and Jack, written almost completely by Liz.   Perhaps dated but
% at least a jumping off point.

The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology. Programming scientific applications is hard, and optimizing them for high performance is even harder.  The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and of the compiler being used. Hardware platforms are continually evolving. In addition, experiments show that the complexity of this  hardware-software system means that performance is difficult to predict fully.  Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is lower (and sometimes significantly lower) than peak.

@@

In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.    [Ugly] Beginning with a MATLAB prototype, we first generate C or Fortran then run it through Pluto~\cite{} to optimize the sequence of loops and then Pluto~\cite{} to generate a collection of organizations of code and to identify the best choice.   Comment on manual parts of this.

@@

The best performance today is obtained through approaches like that taken by PERI~\cite{PERI}, which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists. Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated. We propose to investigate ways to capture this performance tuning expertise in a way that can be integrated into the application. We will thus enable performance portability and ensure that the improvements to the application are long-lasting.

% TODO: Paragraph about source-to-source transformation tools, including PLuTo, Orio

@@

The PIs and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS ~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.  Single routines that carry out more than one linear algebra operation are known as {\t composed} BLAS. As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
bottlenecks of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}  and of the GEMVER kernel examined in Section~\ref{sec:matlab}.   These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies. In the former case, the matrix $A$ is accessed twice while, in the latter, is it accessed only once. Our preliminary results indicate that this loop fusion leads to a routine that delivers 90\% more megaflops than does a pair of calls to the highly optimized Goto BLAS~\cite{Goto:2006fk} {\tt GEMV} routines for large matrix orders.

@@

To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach, illustrated in Figure~\ref{fig:process}.    To begin, we have developed a compiler that converts a MATLAB script into simple C code.   We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously. The result is an automatically generated OpenMP parallel code for multicores.   Finally, we employ the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters and perform an empirical search for selecting the best among the multiple optimized code variants.

Probably need some words about WHY this 3 pronged approach is a good idea.

@@

Figure~\ref{fig:compiler} gives an overview of the compilation process. The MATLAB kernel specification is parsed into a high-level intermediate representation in the form of a dataflow graph.  This dataflow graph is then iteratively processed until all of the implementation choices have been made.  The compilation process consists of three phases: analysis, lowering, and optimization, which are together iterated until all of the implementation decisions have been made.  The graph is then translated into C code.

@@

\subsubsection{Search Space Exploration and Evaluation}
\label{sec:search}

This will need updating depending on how we decide to address this topic. -Jeremy 8/6/08 10:19 AM 
The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq, Cooper:2005kx}.  Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time. Our prototype already performs some pruning: the optimization rewrite rules only apply under certain circumstances.  However, as mentioned above we are sometimes overly aggressive in applying optimizations. To solve this problem we plan to use a combination of analytic methods and empirical methods~\cite{Chen:2007vn}.
%Yotov:2005kx
We plan to use a fast analytic cost function to prune out regions of the search space that will not produce competitive implementations. The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.  Once the search space is narrowed, we will use empirical testing with representative data sets to determine the exact performance of the alternatives and choose the best one.

@@

A graph that cannot be further refined has been reduced to a collection of subgraphs representing loops over scalar operations. The generator outputs a C loop for each subgraph.  The order in which the loops, and scalar operations within the loops, appear in the output is determined by performing a topological sort.

\subsection{PLuTo}
\label{sec:plutonorio}
% http://www.cse.ohio-state.edu/~bondhugu//

PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of (possibly nested) loops.

\subsection{Orio}
\label{sec:orio}

@@

@@

@@

In this section we present performance results for a number of composed BLAS operations. We performed experiments on an dual quad-core 2/88 Xeon with 2.88 GHz, 16GB DDR2 FBDIMM

@@

@@

$y = A' * (A * x)$

\begin{figure}[htp]
\centering
\includegraphics[width=.8\textwidth]{figures/atax.png}
\caption{Wall-clock time (in seconds) for the ATAX operation using different code versions.}
\label{fig:atax}
\end{figure}

@@

\begin{figure}[htb]
\centering
\scriptsize
\begin{verbatim}
for (i=0; i<=A_nrows-1; i++)
  for (j=0; j<=A_ncols-1; j++)
   B[i][j] = A[i][j] + u1[i]*v1[j] + u2[i]*v2[j];

for (i=0; i<=A_ncols-1; i++)
  for (j=0; j<=A_nrows-1; j++)
    x[i] = x[i] + b * B[j][i]*y[j];

for (i=0; i<=A_ncols-1; i++)
  x[i] = x[i] + z[i];

for (i=0; i<=A_nrows-1; i++)
  for (j=0; j<=A_ncols-1; j++)
    w[i] = w[i] + a * B[i][j]*x[j];
\end{verbatim}
\caption{GEMVER C code.}
\label{fig:gemver}
\end{figure}

%--- Sequential: seconds (dynamic arrays) ---
%cmatlab= [0.038, 0.157, 0.352, 0.626, 0.976, 1.406, 1.909, 2.494, 3.143, 3.886]
%orio= [0.029, 0.118, 0.263, 0.480, 0.759, 1.075, 1.447, 1.899, 2.338, 2.978]
%blas= [0.064, 0.265, 0.598, 1.091, 1.711, 2.456, 3.340, 4.345, 5.480, 6.840]
%mkl=  [0.073, 0.147, 0.331, 0.591, 0.923, 1.326, 1.803, 2.350, 2.975, 3.676]
%atlas=[0.039, 0.163, 0.401, 0.796, 1.225, 1.916, 2.582, 3.381, 4.210, 4.756]
%
%--- Parallel: seconds (dynamic arrays) --- N=10,000, cores=1-8
%cmatlab= [0.976, 0.976, 0.976, 0.976, 0.976, 0.976, 0.975, 0.976]
%orio= [0.754, 0.738, 0.666, 0.649, 0.664, 0.664, 0.647, 0.668]

Experiments on Intel Xeon and BGP.

\begin{figure}[htp]
\centering
\includegraphics[width=.8\textwidth]{figures/gemver.png}
\caption{Wall-clock time (in seconds) for the GEMVER operation using different code versions.}
\label{fig:gemver}
\end{figure}

\subsection{GESUMMV}
$y = a * (A * x) + b * (B * x)$

\begin{figure}[htp]
\centering
\includegraphics[width=.8\textwidth]{figures/gesummv.png}

\caption{Wall-clock time (in seconds) for the GESUMMV operation using different code versions.}
\label{fig:gesummv}
\end{figure}
\subsection{BiCG Kernel}

@@

>>>>

% CGO 09: http://www.cgo.org/
% svn co https://svn.mcs.anl.gov/repos/performance/orio/doc/papers/cgo08
% Max 22 pages, double-spaced, 11pt or larger font
% Abstract 300 words: Sep. 12, 2008
% Full paper: Sep. 19, 2008

\documentclass[11pt]{article}

@@

\abstract{

The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology. Programming scientific applications is hard, and optimizing them for high performance is even harder.  The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations. Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.  Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.

@@

The development of high-performance numerical codes is challenging because performance is determined by complex interactions between the algorithm, data structure, programming language, compiler, and computer architecture. Scientists seeking high performance are thus required to master advanced concepts in computer science and carry out intricate programming tasks in addition to managing the scientific content of their work.  They must either invest substantial time in tuning their software or accept low performance. In either case, the productivity of the scientist degrades.

@@

We are working on a combination of the two approaches designed to overcome some of their shortcomings.   This paper described our three-pronged approach toward the development of an automatically tuned library for matrix algebra computations.  In Section~\ref{sec:matlab}, we describe the compiler we have developed that converts a MATLAB script into simple C code. In Section~\ref{sec:orio}, we demonstrate how to employ the empirical performance tuning system Orio in conjunction with the PLuTo tool to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment.  The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism. In Section~\ref{sec:experiments}, we provide performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
%Text in intro repeats abstract some, but that might not be a big deal for a conf paper.

@@

The best performance today is obtained through approaches like that taken by PERI~\cite{PERI} which involves a labor intensive rewrite of portion of the code by a group of performance tools experts in collaboration with application scientists. Maintaining performance of the code as the application and hardware evolves means that the tuning process may need to be repeated. We propose to investigate ways to capture this performance tuning expertise in a way that can be integrated into the application. We will thus enable performance portability and ensure that the improvements to the application are long-lasting.

A number of source or binary transformation tools for performance-improving optimizations exist. LoopTool~\cite{LoopTool}, developed at Rice University, supports annotation-based loop fusion, unroll-and-jam, skewing and tiling.  A new tool, POET~\cite{POET} also supports a number of loop transformations. POET offers a complex template-based syntax for defining transformations in a language-independent manner. Other recent research efforts whose goal, at least in part, is to enable optimizations of source code to be augmented with performance-related information include the X language~\cite{XLanguage} (a macro C-like language for annotating C code), the Broadway~\cite{broadway} compiler,  telescoping languages~\cite{telescopingurl,teleoverview,Ken99}, and various meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}. PLuTo~\cite{Pluto,uday08cc} is a source-to-source transformation tool for optimizing sequences of nested loops. PLuTo employs a polyhedral model of nested loops, where the dynamic instance (iteration) of each statement is viewed as an integer point in a well-defined space, called the statement's polyhedron. Combined with a characterization of data dependences, this representation allows the construction of mathematically correct complex loop transformations. The transformations target both improved cache locality and parallelism.

@@

The authors and other researchers have observed that it can be much more efficient to call a single specialized routine that performs multiple operations than to make successive calls to separate BLAS ~\cite{Ashby:uq,Blackford:2002vn,Stanley:CSD-98-992,baker03blgmres,baker03lgmres,Dennis:2005tg,Howell:2008,gropp01,Vuduc:2003kl}.  Single routines that carry out more than one linear algebra operation are known as {\t composed} BLAS. As an example, consider the pair of matrix-vector products $q  = Ap$, $s  = A^Tr$, where $A$ is a matrix and $p$, $q$, $r$, and $s$ are vectors, that comprise the computational
bottlenecks of the BiConjugate Gradient Method~\cite{Barrett:1994kx,Saad:2003fk}  and of the GEMVER kernel examined in Section~\ref{sec:matlab}.   These two operations can be implemented as a pair of calls to the BLAS routine {\tt GEMV} or they can be rewritten as a composed BLAS consisting of the doubly nested loop encompassing both matrix-vector multiplies. In the former case, the matrix $A$ is accessed twice while, in the latter, is it accessed only once. Our preliminary results indicate that this loop fusion leads to a routine that delivers 90\% more megaflops than does a pair of calls to the highly optimized Goto BLAS~\cite{Goto:2006fk} {\tt GEMV} routines for large matrix orders.

@@

To generate highly optimized code from a MATLAB prototype, we will follow a three-step approach, illustrated in Figure~\ref{fig:process}.    To begin, we have developed a compiler that converts a MATLAB script into simple C code.   We then use the source-to-source automatic parallelization tool  PLuTo\cite{Pluto} to optimize for coarse-grained parallelism and locality simultaneously. The result is an automatically generated code containing transformed loops and OpenMP directives for compiler parallelization.  Finally, we employ the extensible annotation system Orio to generate many tuned versions of the same operation using different optimization parameters and perform an empirical search for selecting the best among the multiple optimized code variants.
%Does this repeat abstract/intro too closely?

%Probably need some words about WHY this 3 pronged approach is a good idea.

@@

We have developed a compiler that translates MATLAB scripts into C code.    Figure~\ref{fig:compiler} gives an overview of the compilation process. The MATLAB kernel specification is parsed into a high-level intermediate representation in the form of a dataflow graph.  This dataflow graph is then iteratively processed until all of the implementation choices have been made.  The compilation process consists of three phases: analysis, lowering, and optimization, which are together iterated until all of the implementation decisions have been made.  The graph is then translated into C code.

@@

@@

A graph that cannot be further refined has been reduced to a collection of subgraphs representing loops over scalar operations. The generator outputs a C loop for each subgraph.  The order in which the loops, and scalar operations within the loops, appear in the output is determined by performing a topological sort
\subsection{Orio}
\label{sec:orio}

@@

% Explain how Pluto is used

% Forward ref. to search approaches

@@

\subsection{Search Space Exploration and Evaluation}
\label{sec:search}

This will need updating depending on how we decide to address this topic. -Jeremy 8/6/08 10:19 AM
% BN: moved to its own section since multiple tools deal with the search space problem.
The set of all possible combinations of implementation and optimization choices is called the search space~\cite{Kisuki:2000uq,Triantafyllis:2003uq, Cooper:2005kx}.  Although the dataflow graphs are quite small in our setting (usually less than 100 nodes), it is still important to prune the search space to avoid an exponential compilation time. Our prototype Matlab compiler already performs some pruning: the optimization rewrite rules only apply under certain circumstances. 
%
% BN: commented out below since it's more future work
%However, as mentioned above we are sometimes overly aggressive in applying optimizations.
%To solve this problem we plan to use a combination of analytic methods and empirical methods~\cite{Chen:2007vn}.
%
Orio also offers a number of different general strategies for narrowing the search space over loop-level optimizations, including random, simplex, and simulated annealing searches.

@@

In this section we present performance results for a number of composed BLAS operations. We performed experiments on an Intel Xeon 2.80 GHz Intel(R) E5462 dual quad-core workstation with 16GB DDR2 FBDIMM RAM.
% Include peak numbers?

@@

BN: working on this

@@

The extended Matlab code for the ATAX operation, $y = A' * (A * x)$,
and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.

\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.36\textwidth}
\footnotesize
\begin{verbatim}
ATAX
in
  A : row matrix, x : vector
out
  y : vector
{
  y = A' * (A * x)
}

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/atax.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the ATAX operation using different code versions.}
\label{fig:atax}
\end{figure}

@@

The extended Matlab code for the GEMVER operation and the corresponding performance results are shown in Figure~\ref{fig:gemver}.

\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.36\textwidth}
\footnotesize
\begin{verbatim}
GEMVER
in
  A : column matrix,
  u1 : vector, u2 : vector,
  v1 : vector, v2 : vector,
  a : scalar, b : scalar,
  y : vector, z : vector
out
  B : column matrix, 
  x : vector, w : vector
{
  B = A + u1 * v1' + u2 * v2'
  x = b * (B' * y) + z
  w = a * (B * x)
}
\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/gemver.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GEMVER operation using different code versions.}
\label{fig:gesummv}
\end{figure}

% Data (Dual quad-core Intel Xeon E5462, 16GB RAM)
%--- Sequential: seconds (dynamic arrays) ---
%cmatlab= [0.038, 0.157, 0.352, 0.626, 0.976, 1.406, 1.909, 2.494, 3.143, 3.886]
%orio= [0.029, 0.118, 0.263, 0.480, 0.759, 1.075, 1.447, 1.899, 2.338, 2.978]
%blas= [0.064, 0.265, 0.598, 1.091, 1.711, 2.456, 3.340, 4.345, 5.480, 6.840]
%mkl=  [0.073, 0.147, 0.331, 0.591, 0.923, 1.326, 1.803, 2.350, 2.975, 3.676]
%atlas=[0.039, 0.163, 0.401, 0.796, 1.225, 1.916, 2.582, 3.381, 4.210, 4.756]
%
%--- Parallel: seconds (dynamic arrays) --- N=10,000, cores=1-8
%cmatlab= [0.976, 0.976, 0.976, 0.976, 0.976, 0.976, 0.975, 0.976]
%orio= [0.754, 0.738, 0.666, 0.649, 0.664, 0.664, 0.647, 0.668]

\subsection{GESUMMV}

The extended Matlab code for the GESUMMV operation, $y = a * (A * x) + b * (B * x)$,
and the corresponding performance results are shown in Figure~\ref{fig:gesummv}.

\begin{figure}[htp]
\centering
\begin{tabular}{cc}
\begin{minipage}[b]{.36\textwidth}
\footnotesize
\begin{verbatim}
GESUMMV
in
  A : row matrix,
  B : row matrix,
  x : vector,
  a : scalar,
  b : scalar
out
  y : vector
{
  y = a * (A * x) + b * (B * x)
}

\end{verbatim}
\end{minipage}
&
\begin{minipage}[b]{.6\textwidth}
\includegraphics[width=\textwidth]{figures/gesummv.png}
\end{minipage}\\
\end{tabular}
\caption{Annotated MATLAB code (left) and wall-clock time in seconds (right) for the GESUMMV operation using different code versions.}
\label{fig:gesummv}
\end{figure}

% BN: ATLAS's dgemv is annihilating everything else, even though it's called twice... wow.
% I did check the result and it was correct
% The Time and MFLOPS data for the above are (with icc):
% (Dual quad-core Intel Xeon E5462, 16GB RAM)
% ops = (4*N*N+3*N)
% --- Parallel: seconds (dynamic arrays) ---
%matlabc= [0.019, 0.075, 0.170, 0.301, 0.471, 0.678, 0.923, 1.205, 1.529, 1.887]
%blas= [0.00006, 0.00018, 0.00039, 0.00060, 0.00090, 0.00126, 0.00168, 0.00216, 0.00270, 0.00330]
%orio= [0.013, 0.035, 0.076, 0.133, 0.210, 0.300, 0.410, 0.533, 0.676, 0.831]
%--- Parallel: Mflops/sec (dynamic arrays) ---
%matlabc= [846.565, 848.425, 849.028, 849.503, 849.690, 849.968, 849.605, 849.946, 847.893, 848.023]
%blas= [271288.136, 347891.304, 367392.857, 421785.832, 480228.091, 497440.415, 508787.800, 516153.226, 524930.741, 530172.300]
%orio= [1231.989, 1846.163, 1885.595, 1918.746, 1907.976, 1920.818, 1910.867, 1920.502, 1915.970, 1924.623]
%
% and gcc behaves similarly, so it's not icc being smart and detecting 2 calls to dgemv
% time: blas= [0.000, 0.000, 0.000, 0.001, 0.001, 0.001, 0.002, 0.002, 0.003, 0.003]
% mflops: blas= [254063.492, 346010.811, 393491.803, 421785.832, 439593.407, 453214.792, 463108.092, 470394.120, 462381.020, 469088.244]

\subsection{BiCG Kernel}

@@

% BN: moved from search space section above
%Yotov:2005kx
We plan to use a fast analytic cost function to prune out regions of the search space that will not produce competitive implementations. The cost function is based on a model of the computer architecture, but does not necessarily account for all the details.  Once the search space is narrowed, we will use empirical testing with representative data sets to determine the exact performance of the alternatives and choose the best one.
